---
title: "MSA2dist Vignette"
author: "Kristian K Ullrich"
date: "`r Sys.Date()`"
abstract: >
    MSA2dist calculates pairwise distances between all sequences of
    a DNAStringSet or a AAStringSet using a custom score matrix and conducts
    codon based analysis. It uses scoring matrices to be used in these pairwise
    distance calculations which can be adapted to any scoring for DNA or AA
    characters. E.g. by using literal distances MSA2dist calcualtes pairwise
    IUPAC distances. DNAStringSet alignments can be analysed as codon alignments
    to look for synonymous and nonsynonymous substitutions (dN/dS) in a
    parallelised fashion using a variety of substitution models. Non-aligned
    coding sequences can be directly used to construct pairwise codon
    alignments (global/local) and calculate dN/dS without any external
    dependencies.
bibliography: bibliography.bib
nocite: '@*'
output: BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{MSA2dist Vignette}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup-packages, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
suppressPackageStartupMessages({
    library(MSA2dist)
    library(Biostrings)
    library(ape)
    library(dplyr)
    library(tidyr)
    library(ggplot2)
    })
```

# Introduction

Calculating pairwise distances of either DNA or AA sequences is a common task
for evolutionary biologist. The distance calculations are either based on
specific nucleotide, codon or amino acid models or on a scoring matrix.

__Note:__ Sequences need to be pre-aligned into so called multiple sequence
alignments (MSA), which can be done with a multitude of existing software.
Just to mention for example [mafft](https://mafft.cbrc.jp/alignment/server/),
[muscle](https://www.ebi.ac.uk/Tools/msa/muscle/) or the R package
[`msa`](https://bioconductor.org/packages/release/bioc/html/msa.html).

The R package [`ape`](https://cran.r-project.org/web/packages/ape/index.html)
for example offers the `ape::dist.dna()` function, which has implemented a
collection of different evolutionary models.
[`MSA2dist`](http://bioconductor.org/
packages/release/bioc/html/MSA2dist.html)
extends the possibility to directly calculate pairwise nucloetide
distances of an `Biostrings::DNAStringSet` object or pairwise amino acid
distances of an `Biostrings::AAStringSet` object. The scoring matrix based
calculations are implemented in `c++` with `RcppThread` to parallelise pairwise
combinations.

It is a non-trivial part to resolve haploid (1n) sequences
from a diploid (2n) individual (aka phasing) to further use the
haploid sequences for distance calculations. To cope with this situation,
`MSA2dist` uses a literal distance [@chang2017whole] which can be directly
applied on `IUPAC` nucleotide ambiguity encoded sequences with the
`dnastring2dist()` function. `IUPAC` sequences can be for example obtained
directly from mapped `BAM` files and the [angsd](http://www.popgen.dk/
angsd/index.php/Fasta) `-doFasta 4` option [@korneliussen2014angsd].

The Grantham's score [@grantham1974amino] attempts to predict the distance
between two amino acids, in an evolutionary sense considering the amino acid
composition, polarity and molecular volume. `MSA2dist` offers with the
`aastring2dist()` function the possibility to obtain pairwise distances of all
sequences in an `Biostrings::AAStringSet` (needs to be pre-aligned). The
resulting distance matrix can be used to calculate neighbor-joining trees via
the R package [`ape`](https://cran.r-project.org/web/packages/ape/index.html).

Calculating synonymous (Ks) and nonsynonymous (Ka) substitutions from coding
sequences and its ratio Ka/Ks can be used as an indicator of selective
pressure acting on a protein. The `dnastring2kaks()` function can be applied on
pre-aligned `Biostrings::DNAStringSet()` objects to calculate these values
either according to [@li1993unbiased] via the R package
[`seqinr`](https://cran.r-project.org/web/packages/seqinr/index.html) or
according to the model of [@nei1986simple].

Further, all codons can be evaluated among the coding sequence alignment and
be plotted to for example protein domains with substitutions or indels with the
`codonmat2xy()` function.

# Installation

To install this package, start R (version "4.1") and enter:

```
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("MSA2dist")
```

# Load MSA2dist

```{r load-data}
# load MSA2dist
library(MSA2dist)
# load example data
data(hiv, package="MSA2dist")
data(iupac, package="MSA2dist")
data(AAMatrix, package="MSA2dist")
data(woodmouse, package="ape")
```

# Sequence Format conversion

To be able to use distance calculation functions from other R packages, like
[`ape`](https://cran.r-project.org/web/packages/ape/index.html) or
[`seqinr`](https://cran.r-project.org/web/packages/seqinr/index.html), it is
necessary to have dedicated sequence format conversion functions. Here, some
examples are shown, how to convert from and to a `Biostrings::DNAStringSet`
object.

`?Biostrings::DNAStringSet()` >>> `?seqinr::as.alignment()`

```{r dnastring2aln}
## define two cds sequences
cds1 <- Biostrings::DNAString("ATGCAACATTGC")
cds2 <- Biostrings::DNAString("ATG---CATTGC")
cds1.cds2.aln <- c(Biostrings::DNAStringSet(cds1),
    Biostrings::DNAStringSet(cds2))
## define names
names(cds1.cds2.aln) <- c("seq1", "seq2")
## convert into alignment
cds1.cds2.aln |> dnastring2aln()
```

`?seqinr::as.alignment()` >>> `?Biostrings::DNAStringSet()`

```{r aln2dnastring}
## convert back into DNAStringSet
cds1.cds2.aln |> dnastring2aln() |> aln2dnastring()
```

`?Biostrings::DNAStringSet()` >>> `?ape::DNAbin()`

```{r dnastring2dnabin}
## convert into alignment
cds1.cds2.aln |> dnastring2dnabin()
```

`?ape::DNAbin()` >>> `?Biostrings::DNAStringSet()`

```{r dnabin2dnastring}
## convert back into DNAStringSet
cds1.cds2.aln |> dnastring2dnabin() |> dnabin2dnastring()
## use woodmouse data
woodmouse |> dnabin2dnastring()
```

`?Biostrings::AAStringSet()` >>> `?seqinr::as.alignment()`

```{r aastring2aln}
## translate cds into aa
aa1.aa2.aln <- cds1.cds2.aln |> cds2aa()
## convert into alignment
aa1.aa2.aln |> aastring2aln()
```

`?seqinr::as.alignment()` >>> `?Biostrings::AAStringSet()`

```{r aln2aastring}
## convert back into AAStringSet
aa1.aa2.aln |> aastring2aln() |> aln2aastring()
```

`?Biostrings::AAStringSet()` >>> `?ape::as.AAbin()`

```{r aastring2aabin}
## convert into AAbin
aa1.aa2.aln |> aastring2aabin()
```

`?ape::as.AAbin()` >>> `?Biostrings::AAStringSet()`

```{r aabin2aastring}
## convert back into AAStringSet
aa1.aa2.aln |> aastring2aabin() |> aabin2aastring()
```

# Frame aware `Biostrings::DNAStringSet` translation (`cds2aa()`)

To be able to translate a coding sequence into amino acids, sometimes the
sequences do not start at the first frame. The `cds2aa` function can take an
alternative codon start site into account (`frame = 1` or `frame = 2` or
`frame = 3`). However, sometimes it is also
necessary that the resulting coding sequence length is a multiple of three.
This can be forced by using the `shorten = TRUE` option.

Simple translation:

```{r cds2aa}
## define two cds sequences
cds1 <- Biostrings::DNAString("ATGCAACATTGC")
cds2 <- Biostrings::DNAString("ATG---CATTGC")
cds1.cds2.aln <- c(Biostrings::DNAStringSet(cds1),
    Biostrings::DNAStringSet(cds2))
## define names
names(cds1.cds2.aln) <- c("seq1", "seq2")
## translate cds into aa
cds1.cds2.aln |> cds2aa()
aa1.aa2.aln <- cds1.cds2.aln |> cds2aa()
```

Translation keeping multiple of three sequence length:

```{r cds2aa-shorten}
## translate cds into aa using frame = 2
## result is empty due to not multiple of three
cds1.cds2.aln |> cds2aa(frame=2)
## translate cds into aa using frame = 2 and shorten = TRUE
cds1.cds2.aln |> cds2aa(frame=2, shorten=TRUE)
## translate cds into aa using frame = 3 and shorten = TRUE
cds1.cds2.aln |> cds2aa(frame=3, shorten=TRUE)
## use woodmouse data
woodmouse |> dnabin2dnastring() |> cds2aa(shorten=TRUE)
```

Translation using alternative genetic code:

As you can see from the above example, the initial amino acids `I` will change
into `M` due to the mitochondrial translation code and also some `*` stop
codons will change into a `W` amino acid.

```{r cds2aa-genetic}
## alternative genetic code
## use woodmouse data
woodmouse |> dnabin2dnastring() |> cds2aa(shorten=TRUE,
    genetic.code=Biostrings::getGeneticCode("2"))
```

# Pairwise sequence comparison

## Calculate pairwise AA distances (`aastring2dist()`)

### Grantham's distance

```{r aastring2dist}
## calculate pairwise AA distances based on Grantham's distance
aa.dist <- hiv |> cds2aa() |> aastring2dist(score=granthamMatrix())
## obtain distances
head(aa.dist$distSTRING)
## obtain pairwise sites used
head(aa.dist$sitesUsed)
```

```{r aastring2dist-plot}
## create and plot bionj tree
aa.dist.bionj <- ape::bionj(as.dist(aa.dist$distSTRING))
plot(aa.dist.bionj)
```

To use a different score matrix, here as an example the `AAMatrix` from the
R package [`alakazam`](https://cran.r-project.org/web/packages/ape/index.html)
is used:

```{r aastring2dist-matrix}
## use AAMatrix data
head(AAMatrix)
aa.dist.AAMatrix <- hiv |> cds2aa() |> aastring2dist(score=AAMatrix)
head(aa.dist.AAMatrix$distSTRING)
```

## Calculate pairwise DNA distances (`dnastring2dist()`)

### `ape::dist.dna` models

```{r dnastring2dist}
## use hiv data
dna.dist <- hiv |> dnastring2dist(model="K80")
## obtain distances
head(dna.dist$distSTRING)
## obtain pairwise sites used
head(dna.dist$sitesUsed)
```

```{r dnastring2dist-bionj}
## create and plot bionj tree
dna.dist.bionj <- ape::bionj(as.dist(dna.dist$distSTRING))
```

It is also possible to compare the amino acid and nucleotide based trees:

```{r treecompare}
## creation of the association matrix:
association <- cbind(aa.dist.bionj$tip.label, aa.dist.bionj$tip.label)
## cophyloplot
ape::cophyloplot(aa.dist.bionj,
    dna.dist.bionj,
    assoc=association,
    length.line=4,
    space=28,
    gap=3,
    rotate=FALSE)
```

### `IUPAC` distance 

```{r dnastring2dist-IUPAC}
## use hiv data
hiv.dist.iupac <- head(hiv |> dnastring2dist(model="IUPAC"))
head(hiv.dist.iupac$distSTRING)
## run multi-threaded
system.time(hiv |> dnastring2dist(model="IUPAC", threads=1))
system.time(hiv |> dnastring2dist(model="IUPAC", threads=2))
```

Woodmouse data example:

```{r dnastring2dist-woodmouse}
## use woodmouse data
woodmouse.dist <- woodmouse |> dnabin2dnastring() |> dnastring2dist()
head(woodmouse.dist$distSTRING)
```

# Coding sequences

## Calculating synonymous and nonsynonymous substitutions (`dnastring2kaks()`)

```{r dnastring2kaks-hiv}
## use hiv data
## model Li
head(hiv |> dnastring2kaks(model="Li"))
## model NG86
head(hiv |> dnastring2kaks(model="NG86", threads=1))
```

## Using any model from KaKs_Calculator 2.0

models to choose from KaKs_Calculator 2.0 [@wang2010kaks_calculator] are:

- MS [@wang2010kaks_calculator]
- MA [@wang2010kaks_calculator]
- GNG [@wang2010kaks_calculator]
- GLWL [@wang2010kaks_calculator]
- GLPB [@wang2010kaks_calculator]
- GMLWL [@wang2010kaks_calculator]
- GMLPB [@wang2010kaks_calculator]
- GYN [@wang2010kaks_calculator]
- LWL [@li1985new]
- LPB [@li1993unbiased, @pamilo1993evolution]
- NG [@nei1986simple]
- MLPB [@tzeng2004comparison]
- GY [@goldman1994codon]
- YN [@yang2000codon]
- MYN [@zhang2006computing, @wang2009gamma]

```{r dnastring2kaks-YN}
## use hiv data
## model MYN
head(hiv |> dnastring2kaks(model="MYN"))
## model YN
head(hiv |> dnastring2kaks(model="YN", threads=1))
```

## Using indices to calculate Ka/Ks

```{r indices2kaks}
## use hiv data
idx <- list(c(2, 3), c(5,7,9))
## model MYN
head(hiv |> indices2kaks(idx, model="MYN"))
## model YN
head(hiv |> indices2kaks(idx, model="YN", threads=1))
```

## Codon comparison

As an example for the codon comparison data from the Human Immunodeficiency
Virus Type 1 is used [@ganeshan1997human], [@yang2000codon].

The window plots are constructed with the R package
[`ggplot2`](https://cran.r-project.org/web/packages/ggplot2/index.html).

### Create codon matrix (`dnastring2codonmat()`)

```{r dnastring2codonmat}
## define two cds sequences
cds1 <- Biostrings::DNAString("ATGCAACATTGC")
cds2 <- Biostrings::DNAString("ATG---CATTGC")
cds1.cds2.aln <- c(Biostrings::DNAStringSet(cds1),
    Biostrings::DNAStringSet(cds2))
## convert into alignment
cds1.cds2.aln |> dnastring2codonmat()
```

Like the `cds2aa()` function, also the `dnastring2codonmat()` function is
implemented to handle different frames.

```{r dnastring2codonmat-frames}
## use frame 2 and shorten to circumvent multiple of three error
cds1 <- Biostrings::DNAString("-ATGCAACATTGC-")
cds2 <- Biostrings::DNAString("-ATG---CATTGC-")
cds1.cds2.aln <- c(Biostrings::DNAStringSet(cds1),
    Biostrings::DNAStringSet(cds2))
cds1.cds2.aln |> dnastring2codonmat(frame=2, shorten=TRUE)
```

### Calculate average behavior of each codon (`codonmat2xy()`)

```{r codonmat2xy}
## use hiv data
## calculate average behavior
hiv.xy <- hiv |> dnastring2codonmat() |> codonmat2xy()
```

### Plot average behavior of each codon

```{r codonmat2xy-plot}
print(hiv.xy |> dplyr::select(Codon,SynMean,NonSynMean,IndelMean) |>
    tidyr::gather(variable, values, -Codon) |>
    ggplot2::ggplot(aes(x=Codon, y=values)) + 
    ggplot2::geom_line(aes(colour=factor(variable))) + 
    ggplot2::geom_point(aes(colour=factor(variable))) + 
    ggplot2::ggtitle("HIV-1 sample 136 patient 1 from
        Sweden envelope glycoprotein (env) gene"))
```

# Dxy and FST

## Dxy and FST from pre-calculated distance matrix (`dist2dxy()`)

Pairwise genetic distances are already available as a full distance matrix.
This allows summarisation of within- and between-population divergence
using full distributional information.

For each population pair, the function reports not only mean values,
but also distributional summaries of pairwise distances, including:

- mean
- median
- minimum
- maximum
- variance
- mean number of aligned sites used

This provides a full empirical description of genetic divergence
structure across populations including additional statistics such
as Gmin, RND, and RNDmin when outgroup information is provided.

```{r dist2dxy}
## create poplist
poplist <- list(FRA = grep("Mmd.FRA", names(iupac)),
    GER = grep("Mmd.GER", names(iupac)),
    IRA = grep("Mmd.IRA", names(iupac)),
    AFG = grep("Mmm.AFG", names(iupac)))
iupac <- iupac |> addpop2string(poplist)
## using popx: FRA, popy: IRA, popout: AFG
dxy <- iupac |> dnastring2dist() |> dist2dxy(pop=popinteger(iupac),
    popx="FRA", popy="IRA", popout="AFG")
dxy$pi_within
dxy$pi_between
```

## Dxy and FST directly from `Biostrings::DNAStringSet` (`dnastring2dxy()`)

When starting from raw sequence data, pairwise distances are computed
on-the-fly from aligned sequences using site-level information
(identity, mismatch, ambiguity handling, or substitution model
depending on settings).

Because distances are not pre-stored as a full matrix,
the output focuses on expected values of divergence,
i.e. population-level means derived from site-based aggregation.

As a result, the primary outputs are mean-based estimators of:

- within-population diversity (π)
- between-population divergence (Dxy)
- derived FST

rather than full distributional summaries of pairwise distances.

```{r dnastring2dxy}
## create poplist
poplist <- list(FRA = grep("Mmd.FRA", names(iupac)),
    GER = grep("Mmd.GER", names(iupac)),
    IRA = grep("Mmd.IRA", names(iupac)),
    AFG = grep("Mmm.AFG", names(iupac)))
iupac <- iupac |> addpop2string(poplist)
dxy <- iupac |> dnastring2dxy(pop=popinteger(iupac), model="IUPAC")
dxy$dxy
dxy$fst
```

# References

<div id="refs"></div>

# Session Info

```{r sessionInfo, echo=TRUE}
sessionInfo()
```
